Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-06 talk/icooolps2009/paper.tex
Lots of fixes here and there.

\newboolean{showcomments}
\setboolean{showcomments}{true}
\ifthenelse{\boolean{showcomments}}
  {\newcommand{\nb}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
   }
   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
  }
  {\newcommand{\nb}[2]{}
   \newcommand{\version}{}
  }

@@

@@

\begin{abstract}
In this paper we describe the ongoing research in the PyPy project to write a
JIT compiler that is automatically adapted to various languages, given an
interpreter for that language. This is achieved with the help of a slightly
adapted tracing JIT compiler in combination with some hints by the author of the
interpreter.  XXX

@@

XXX write somewhere that one problem of using tracing JITs for dynamic languages
is that dynamic languages have very complex bytecodes

@@

Dynamic languages, rise in popularity, bla bla XXX

One of the often-cited drawbacks of dynamic languages is the performance
penalties they impose. Typically they are slower than statically typed languages
\cite{XXX}. Even though there has been a lot of research into improving the
performance of dynamic languages \cite{XXX}, those techniques are not as widely
used as one would expect. Many dynamic language implementations use completely
straightforward bytecode-interpreters without any advanced implementation
techniques like just-in-time compilation. There are a number of reasons for
this. Most of them boil down to the inherent complexities of using compilation.
Interpreters are simple to implement, understand, extend and port whereas writing a
just-in-time compiler is an error-prone task that is made even harder by the
dynamic features of a language.

@@

The contributions of this paper are:
\begin{itemize}
\item Techniques for improving the generated code when applying a tracing JIT to
an interpreter
\item 
\end{itemize}

@@

Tracing JITs are built on the following basic assumptions:

\begin{itemize}
 \item programs spend most of their runtime in loops
 \item several iterations of the same loop are likely to take similar code paths
\end{itemize}

@@

Typically, programs executed by a tracing VMs goes through various phases:

\begin{itemize}
\item Interpretation/profiling
\item Tracing
\item Code generation
\item Execution of the generated code
\end{itemize}

@@

Such a history is called a \emph{trace}: it is a sequential list of
operations, together with their actual operands and results.  By examining the
trace, it is possible to produce highly efficient machine code by emitting
only the operations needed.  Being sequential, the trace represents only one
of the many possible paths through the code. To ensure correctness, the trace
contains a \emph{guard} at every possible point where the path could have
followed another direction, for example conditions or indirect/virtual
calls.  When emitting the machine code, every guard is turned into a quick check
to guarantee that the path we are executing is still valid.  If a guard fails,
we immediately quit from the machine code and continue the execution by falling
ways.

During tracing, the trace is repeatedly
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the interpreted program that the tracing interpreter is running. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
starting from the second iteration of the loop,
as it represents exactly the loop that was being interpreted so far.

This process assumes that the path through the loop that was traced is a
"typical" example of possible paths (which is statistically likely). Of course
it is possible that later another path through the loop is taken, therefore the
machine code will contain \emph{guards}, which check that the path is still the same.
If a guard fails during execution of the machine code, the machine code is left
and execution falls back to using interpretation (there are more complex
mechanisms in place to still produce more code for the cases of guard failures,
but they are of no importance for this paper).

It is important to understand when the tracer considers a loop in the trace to
be closed. This happens when the \emph{position key} is the same as at an earlier
point. The position key describes the position of the execution of the program,
e.g. usually contains things like the function currently being executed and the
program counter position of the tracing interpreter. The tracing interpreter
does not need to check all the time whether the position key already occurred
earlier, but only at instructions that are able to change the position key
to an earlier value, e.g. a backward branch instruction. Note that this is
already the second place where backward branches are treated specially: during
interpretation they are the place where the profiling is performed and where
tracing is started or already existing assembler code entered; during tracing
they are the place where the check for a closed loop is performed.

Let's look at a small example. Take the following (slightly contrived) RPython
code:
{\small
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result
\end{verbatim}
}
To trace this, a bytecode form of these functions needs to be introduced that
the tracer understands. The tracer interprets a bytecode that is an encoding of
the intermediate representation of PyPy's translation toolchain after type
inference has been performed and Python-specifics have been made explicit. At
first those functions will be interpreted, but after a while, profiling shows
that the \texttt{while} loop in \texttt{strange\_sum} is executed often.  The
tracing JIT will then start to trace the execution of that loop.  The trace would
look as follows:
{\small
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}
}
The operations in this sequence are operations of the mentioned intermediate
representation (e.g. note that the generic modulo and equality operations in the
function above have been recognized to always work on integers and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed, is in SSA-form \cite{XXX} and ends with a jump
to its own beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} was inlined into the trace. Note that the trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be turned into machine code and executed.

@@

The tracing JIT of the PyPy project is atypical in that it is not applied to the
user program, but to the interpreter running the user program. In this section
we will explore what problems this brings, and how to solve them (at least
partially). This means that there are two interpreters involved, and we need
terminology to distinguish them. On the one hand, there is the interpreter that
the tracing JIT uses to perform tracing. This we will call the \emph{tracing
interpreter}. On the other hand, there is the interpreter that is running the
users programs, which we will call the \emph{language interpreter}. In the
following, we will assume that the language interpreter is bytecode-based. The
program that the language interpreter executes we will call the \emph{user
program} (from the point of view of a VM author, the "user" is a programmer
using the VM).

@@

\fijal{I find following paragraph out of scope and completely confusing, we
should instead simply state that we unroll the loop, how we do that and
why we do that. Completely ignore aspect of an interpreter loop I suppose,
because everything previously keeps talking about can\_enter\_jit that closes
loop being available at jump back bytecodes}
A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The most important hot loop
of the language interpreter is its bytecode dispatch loop (for many simple
interpreters it is also the only hot loops).  Tracing one iteration of this
loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is just wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
\begin{figure}
\input{code/tlr-paper.py}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}
\begin{figure}
{\small
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

@@

    MOV_R_A     2   # return res
    RETURN_A
\end{verbatim}
}
\caption{Example bytecode: Compute the square of the accumulator}
\label{fig:square}
\end{figure}

\fijal{This paragraph should go away as well}
Let's look at an example. Figure \ref{fig:tlr-basic} shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator. The
\texttt{bytecode} argument is a string of bytes and all register and the
accumulator are integers. A simple program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
Because of the guard on \texttt{opcode0}, the code compiled from this trace will
be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
the other operations the guard will fail, which will mean that performance is
probably not improved at all.

\begin{figure}
\input{code/normal-tracing.txt}
\caption{Trace when executing the \texttt{DECR\_A} opcode}
\label{fig:trace-normal}
\end{figure}

To improve this situation, the tracing JIT could trace the execution of several
opcodes, thus effectively unrolling the bytecode dispatch loop. Ideally, the
bytecode dispatch loop should be unrolled exactly so much, that the unrolled version
corresponds to \emph{user loop}. User loops
occur when the program counter of the language interpreter has the
same value several times. This program counter is typically stored in one or several
variables in the language interpreter, for example the bytecode object of the
currently executed function of the user program and the position of the current
bytecode within that.  In the example above, the program counter is represented by 
the \texttt{bytecode} and \texttt{pair} variables.

@@

The condition for reusing already existing machine code needs to be adapted to
this new situation. In a classical tracing JIT there is at most one piece of
assembler code per loop of the jitted program, which in our case is the language
interpreter. When applying the tracing JIT to the language interpreter as
described so far, \emph{all} pieces of assembler code correspond to the bytecode
dispatch loop of the language interpreter. They correspond to different
unrollings and paths of that loop though. To figure out which of them to use
when trying to enter assembler code again, the program counter of the language
interpreter needs to be checked. If it corresponds to the position key of one of
the pieces of assembler code, then this assembler code can be entered. This
check again only needs to be performed at the backward branches of the language
interpreter.

@@

\begin{figure}
\input{code/tlr-paper-full.py}
\caption{Simple bytecode interpreter with hints applied}
\label{fig:tlr-full}
\end{figure}

Let's look at which hints would need to be applied to the example interpreter
from Figure \ref{fig:tlr-basic}. The basic thing needed to apply hints is a
subclass of \texttt{JitDriver} that lists all the variables of the bytecode
loop. The variables are classified into two groups, red variables and green
variables. The green variables are those that the tracing JIT should consider to
be part of the program counter of the language interpreter. In the case of the
example, the \texttt{pc} variable is obviously part of the program counter.
However, the \texttt{bytecode} variable is also counted as green, since the
\texttt{pc} variable is meaningless without the knowledge of which bytecode
string is currently being interpreted. All other variables are red.

@@

\begin{figure}
\input{code/no-green-folding.txt}
\caption{Trace when executing the Square function of Figure \ref{fig:square},
with the corresponding bytecodes as comments.}
\label{fig:trace-no-green-folding}
\end{figure}

XXX summarize at which points the tracing interpreter needed changing
XXX all changes only to the position key and when to enter/leave the tracer!
XXX tracing remains essentially the same

@@

The simple insight how to greatly improve the situation is that most of the
operations in the trace are actually concerned with manipulating the
bytecode and the program counter. Those are stored in variables that are part of
the position key (they are "green"), that means that the tracer checks that they
are some fixed value at the beginning of the loop. In the example the check
would be that the \texttt{bytecode} variable is the bytecode string
corresponding to the square function and that the \texttt{pc} variable is
\texttt{4}. Therefore it is possible to constant-fold computations on them away,
as long as the operations are side-effect free. Since strings are immutable in
RPython, it is possible to constant-fold the \texttt{strgetitem} operation. The
\texttt{int\_add} are additions of the green variable \texttt{pc} and a true
constant, so they can be folded away as well.

With this optimization enabled, the trace looks as in Figure
\ref{fig:trace-full}. Now a lot of the language interpreter is actually gone
from the trace and what is left corresponds very closely to the loop of the
square function. The only vestige of the language interpreter is the fact that
the register list is still used to store the state of the computation. This
could be removed by some other optimization, but is maybe not really all that
bad anyway (in fact we have an experimental optimization that does exactly that,
but it is not finished).

\anto{XXX I propose to show also the trace with the malloc removal enabled, as it
  is much nicer to see. Maybe we can say that the experimental optimization we
  are working on would generate this and that} \cfbolz{This example is not about
  mallocs! There are no allocations in the loop. The fix would be to use
  maciek's lazy list stuff (or whatever it's called) which is disabled at the
  moment}

\begin{figure}
\input{code/full.txt}
\caption{Trace when executing the Square function of Figure \ref{fig:square},
with the corresponding opcodes as comments. The constant-folding of operations
on green variables is enabled.}
\label{fig:trace-full}
\end{figure}

Once we get this highly optimized trace, we can pass it to the \emph{JIT
backend}, which generates the correspondent machine code.

@@

If the JIT is enabled, things are more interesting. At the moment the JIT can
only be enabled when translating the interpreter to C, but we hope to lift that
restriction in the future. A classical tracing JIT will
interpret the program it is running until a common loop is identified, at which
point tracing and ultimately assembler generation starts. The tracing JIT in
PyPy is operating on the language interpreter, which is written in RPython. But
RPython programs are statically translatable to C anyway. This means that interpreting the
language interpreter before a common loop is found is clearly not desirable,
since the overhead of this double-interpretation would be significantly too big
to be practical.

What is done instead is that the language interpreter keeps running as a C
program, until a common loop in the user program is found. To identify loops the
C version of the language interpreter is generated in such a way that at the
place that corresponds to the \texttt{can\_enter\_jit} hint profiling is
performed using the program counter of the language interpreter. Apart from this
bit of profiling, the language interpreter behaves in just the same way as
without a JIT.

When a hot user loop is identified, tracing is started. The
tracing interpreter is invoked to start tracing the language interpreter that is
running the user program. Of course the tracing interpreter cannot actually
trace the execution of the C representation of the language interpreter. Instead
it takes the state of the execution of the language interpreter and starts
tracing using a bytecode representation of the language interpreter. That means
there are two "versions" of the language interpreter embedded in the final
executable of the VM: on the one hand it is there as executable machine code, on
the other hand as bytecode for the tracing interpreter. It also means that
tracing is costly as it incurs exactly a double interpretation overhead.

@@

\textbf{Trace Trees:} This paper ignored the problem of guards that fail in a
large percentage of cases because there are several equally likely paths through
a loop. Just falling back to interpretation in this case is not practicable.
Therefore, if we find a guard that fails often enough, we start tracing from
there and produce efficient machine code for that case, instead of alwayas
falling back to interpretation.

@@

In this section we try to evaluate the work done so far by looking at some
benchmark numbers. Since the work is not finished, these benchmarks can only be
preliminary. All benchmarking was done on an otherwise idle machine with a 1.4
GHz Pentium M processor and 1GiB RAM, using Linux 2.6.27.

@@

\begin{enumerate}
\item The interpreter translated to C without any JIT inserted at all.
\item The tracing JIT is enabled, but no interpreter-specific
hints are applied. This corresponds to the trace in Figure
\ref{fig:trace-normal}. The time includes the time it takes to trace and the
production of the machine code, as well as the fallback interpreter to leave the
machine code. The threshold when to consider a loop to be hot is 40 iterations.
\item The hints as in Figure \ref{fig:tlr-full} are applied, which means the loop of
the square function is reflected in the trace. Constant folding of green
variables is disabled though. This corresponds to the trace in Figure
\ref{fig:trace-no-green-folding}. XXX
\item Same as before, but with constant folding enabled. This corresponds to the
trace in Figure \ref{fig:trace-full}. This speeds up the square function nicely,
making it about six times faster than the pure interpreter.
\item Same as before, but with the threshold set so high that the tracer is
never invoked. This measures the overhead of the profiling. For this interpreter
the overhead seems rather large, with 50\% slowdown due to profiling. This is
because the example interpreter needs to do one hash table lookup per loop
iteration. For larger interpreters (e.g. the Python one) it seems likely that
the overhead is less significant, given that many operations in Python need
hash-table lookups themselves.
\item Runs the whole computation on the tracing interpreter for estimating the
involved overheads of tracing. The trace is not actually recorded (which would be a
memory problem), so in reality the number is even higher. Due to the double
interpretation, the overhead is huge. It remains to be seen whether that will be
a problem for practical interpreters.
\item For comparison, the time of running the interpreter on top of CPython
(version 2.5.2).
\end{enumerate}

\begin{figure}
\noindent
\begin{tabular}{|l|r|}
\hline
 &ratio\tabularnewline
\hline
Interpreter compiled to C, no JIT &1\tabularnewline \hline
Normal Trace Compilation &1.20\tabularnewline \hline
Unfolding of Language Interpreter Loop &XXX\tabularnewline \hline
Full Optimizations &0.17\tabularnewline \hline
Profile Overhead &1.51\tabularnewline \hline
Interpreter run by Tracing Interpreter &860.20\tabularnewline \hline
Interpreter run by CPython &256.17\tabularnewline \hline
\end{tabular}
\label{fig:bench1}
\caption{Benchmark results of example interpreter computing the square of
46340}
\end{figure}

%- benchmarks
%    - running example
%    - gameboy?

@@

Applying a trace-based optimizer to an interpreter and adding hints to help the
tracer produce better results has been tried before in the context of the DynamoRIO
project \cite{sullivan_dynamic_2003}. This work is conceptually very close to
ours. They achieve the same unrolling of the interpreter loop so that the
unrolled version corresponds to the loops in the user program. However the
approach is greatly hindered by the fact that they trace on the machine code
level and thus have no high-level information available about the interpreter.
This makes it necessary to add quite a large number of hints, because at the
assembler level it is not really visible anymore that e.g. a bytecode string is
really immutable. Also more advanced optimizations like allocation removal would
not be possible with that approach.

>>>>

\newboolean{showcomments}
\setboolean{showcomments}{false}
\ifthenelse{\boolean{showcomments}}
  {\newcommand{\nb}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
   }
   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
  }
  {\newcommand{\nb}[2]{}
   \newcommand{\version}{}
  }

@@

\textfloatsep 12pt plus 2pt minus 4pt

@@

\begin{abstract}
We present techniques for improving the results when a tracing JIT compiler is
applied to an interpreter. An unmodified tracing JIT performs not as well as one
would hope when the compiled program is itself a bytecode interpreter. We
examine why that is the case, and how matters can be improved by adding hints to
the interpreter, that help the tracing JIT to improve the results. We evaluate
the techniques by using them both on a small example as well as on a full Python
interpreter. This work has been done in the context of the PyPy project.

@@

@@

Dynamic languages have seen a steady rise in popularity in recent years.
JavaScript is increasingly being used to implement full-scale applications
running in browser, whereas other dynamic languages (such as Ruby, Perl, Python,
PHP) are used for the server side of many web sites, as well as in areas
unrelated to the web.

One of the often-cited drawbacks of dynamic languages is the performance
penalties they impose. Typically they are slower than statically typed
languages. Even though there has been a lot of research into improving the
performance of dynamic languages (in the SELF project, to name just one example
\cite{XXX}), those techniques are not as widely
used as one would expect. Many dynamic language implementations use completely
straightforward bytecode-interpreters without any advanced implementation
techniques like just-in-time compilation. There are a number of reasons for
this. Most of them boil down to the inherent complexities of using compilation.
Interpreters are simple to implement, understand, extend and port whereas writing a
just-in-time compiler is an error-prone task that is made even harder by the
dynamic features of a language.

@@

The contributions of this paper are:
\vspace{-0.3cm}
\begin{zitemize}
\item Applying a tracing JIT compiler to an interpreter.
\item Finding techniques for improving the generated code.
\item Integrating
\end{zitemize}

@@

Tracing JITs are built on the following basic assumptions:
\vspace{-0.3cm}
\begin{zitemize}
 \item programs spend most of their runtime in loops
 \item several iterations of the same loop are likely to take similar code paths
\end{zitemize}

@@

Typically, programs executed by a tracing VMs goes through various phases:
\vspace{-0.3cm}
\begin{zitemize}
\item Interpretation/profiling
\item Tracing
\item Code generation
\item Execution of the generated code
\end{zitemize}

@@

Such a history is called a \emph{trace}: it is a sequential list of
operations, together with their actual operands and results.  By examining the
trace, it is possible to produce highly efficient machine code by emitting
only the operations needed.  Being sequential, the trace represents only one
of the many possible paths through the code. To ensure correctness, the trace
contains a \emph{guard} at every possible point where the path could have
followed another direction, for example conditions or indirect/virtual
calls.  When emitting the machine code, every guard is turned into a quick check
to guarantee that the path we are executing is still valid.  If a guard fails,
we immediately quit from the machine code and continue the execution by falling
back to interpretation.

During tracing, the trace is repeatedly
checked whether the interpreter is at a position in the program that it had seen
earlier in the trace. If this happens, the trace recorded corresponds to a loop
in the interpreted program. At this point, this loop
is turned into machine code by taking the trace and making machine code versions
of all the operations in it. The machine code can then be immediately executed,
starting from the next iteration of the loop, as the machine code represents
exactly the loop that was being interpreted so far.

This process assumes that the path through the loop that was traced is a
"typical" example of possible paths (which is statistically likely). Of course
it is possible that later another path through the loop is taken, in which case
one of the guards that were put into the machine code will fail. There are more
complex mechanisms in place to still produce more code for the cases of guard
failures \cite{XXX}, but they are orthogonal to the issues discussed in this
paper.

It is important to understand how the tracer recognizes that the trace it
recorded so far corresponds to a loop.
This happens when the \emph{position key} is the same as at an earlier
point. The position key describes the position of the execution of the program,
e.g. usually contains things like the function currently being executed and the
program counter position of the tracing interpreter. The tracing interpreter
does not need to check all the time whether the position key already occurred
earlier, but only at instructions that are able to change the position key
to an earlier value, e.g. a backward branch instruction. Note that this is
already the second place where backward branches are treated specially: during
interpretation they are the place where the profiling is performed and where
tracing is started or already existing assembler code entered; during tracing
they are the place where the check for a closed loop is performed.

Let's look at a small example. Take the following (slightly contrived) RPython
code:
{\small
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result
\end{verbatim}
}
\vspace{-0.4cm}
To trace this, a bytecode form of these functions needs to be introduced that
the tracer understands. The tracer interprets a bytecode that is an encoding of
the intermediate representation of PyPy's translation toolchain after type
inference has been performed and Python-specifics have been made explicit. At
first those functions will be interpreted, but after a while, profiling shows
that the \texttt{while} loop in \texttt{strange\_sum} is executed often.  The
tracing JIT will then start to trace the execution of that loop.  The trace would
look as follows:
{\small
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}
}
\vspace{-0.4cm}
The operations in this sequence are operations of the mentioned intermediate
representation (e.g. note that the generic modulo and equality operations in the
function above have been recognized to always take integers as arguments and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed, is in SSA-form \cite{cytron_efficiently_1991} and ends with a jump
to its own beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} was inlined into the trace. Note that the trace
contains only the hot \texttt{else} case of the \texttt{if} test in \texttt{f},
while the other branch is implemented via a guard failure. This trace can then
be turned into machine code and executed.

@@

The tracing JIT of the PyPy project is atypical in that it is not applied to the
user program, but to the interpreter running the user program. In this section
we will explore what problems this brings, and how to solve them (at least
partially). This means that there are two interpreters involved, and we need
terminology to distinguish them. On the one hand, there is the interpreter that
the tracing JIT uses to perform tracing. This we will call the \emph{tracing
interpreter}. On the other hand, there is the interpreter that is running the
user's programs, which we will call the \emph{language interpreter}. In the
following, we will assume that the language interpreter is bytecode-based. The
program that the language interpreter executes we will call the \emph{user
program} (from the point of view of a VM author, the "user" is a programmer
using the VM).

@@

\fijal{I find following paragraph out of scope and completely confusing, we
should instead simply state that we unroll the loop, how we do that and
why we do that. Completely ignore aspect of an interpreter loop I suppose,
because everything previously keeps talking about can\_enter\_jit that closes
loop being available at jump back bytecodes}
A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The most important hot loop
of the language interpreter is its bytecode dispatch loop (for many simple
interpreters it is also the only hot loop).  Tracing one iteration of this
loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is just wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
\begin{figure}
\input{code/tlr-paper.py}
\vspace{-0.4cm}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}
\begin{figure}
{\small
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

@@

    MOV_R_A     2   # return res
    RETURN_A
\end{verbatim}
}
\vspace{-0.4cm}
\caption{Example bytecode: Compute the square of the accumulator}
\label{fig:square}
\end{figure}

\fijal{This paragraph should go away as well}
Let's look at an example. Figure \ref{fig:tlr-basic} shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator. The
\texttt{bytecode} argument is a string of bytes, all register and the
accumulator are integers. A program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
Because of the guard on \texttt{opcode0}, the code compiled from this trace will
be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
the other operations the guard will fail, which will mean that performance is
probably not improved at all.

\begin{figure}
\input{code/normal-tracing.txt}
\vspace{-0.4cm}
\caption{Trace when executing the \texttt{DECR\_A} opcode}
\label{fig:trace-normal}
\end{figure}

To improve this situation, the tracing JIT could trace the execution of several
opcodes, thus effectively unrolling the bytecode dispatch loop. Ideally, the
bytecode dispatch loop should be unrolled exactly so much, that the unrolled version
corresponds to a \emph{user loop}. User loops
occur when the program counter of the \emph{language interpreter} has the
same value several times. This program counter is typically stored in one or several
variables in the language interpreter, for example the bytecode object of the
currently executed function of the user program and the position of the current
bytecode within that.  In the example above, the program counter is represented by 
the \texttt{bytecode} and \texttt{pair} variables.

@@

The condition for reusing already existing machine code needs to be adapted to
this new situation. In a classical tracing JIT there is at most one piece of
assembler code per loop of the jitted program, which in our case is the language
interpreter. When applying the tracing JIT to the language interpreter as
described so far, \emph{all} pieces of assembler code correspond to the bytecode
dispatch loop of the language interpreter. They correspond to different
unrollings and paths through that loop though. To figure out which of them to use
when trying to enter assembler code again, the program counter of the language
interpreter needs to be checked. If it corresponds to the position key of one of
the pieces of assembler code, then this assembler code can be entered. This
check again only needs to be performed at the backward branches of the language
interpreter.

@@

\begin{figure}
\input{code/tlr-paper-full.py}
\vspace{-0.4cm}
\caption{Simple bytecode interpreter with hints applied}
\label{fig:tlr-full}
\end{figure}

Let's look at how hints would need to be applied to the example interpreter
from Figure \ref{fig:tlr-basic}. The basic thing needed to apply hints is a
subclass of \texttt{JitDriver} that lists all the variables of the bytecode
loop. The variables are classified into two groups, red variables and green
variables. The green variables are those that the tracing JIT should consider to
be part of the program counter of the language interpreter. In the case of the
example, the \texttt{pc} variable is obviously part of the program counter.
However, the \texttt{bytecode} variable is also counted as green, since the
\texttt{pc} variable is meaningless without the knowledge of which bytecode
string is currently being interpreted. All other variables are red.

@@

\begin{figure}
\input{code/no-green-folding.txt}
\vspace{-0.4cm}
\caption{Trace when executing the Square function of Figure \ref{fig:square},
with the corresponding bytecodes as comments.}
\label{fig:trace-no-green-folding}
\end{figure}

@@

The simple insight how to greatly improve the situation is that most of the
operations in the trace are actually concerned with manipulating the
bytecode and the program counter. Those are stored in variables that are part of
the position key (they are "green"), that means that the tracer checks that they
are some fixed value at the beginning of the loop. In the example the check
would be that the \texttt{bytecode} variable is the bytecode string
corresponding to the square function and that the \texttt{pc} variable is
\texttt{4}. Therefore it is possible to constant-fold computations on them away,
as long as the operations are side-effect free. Since strings are immutable in
RPython, it is possible to constant-fold the \texttt{strgetitem} operation. The
\texttt{int\_add} are additions of the green variable \texttt{pc} and a constant
number, so they can be folded away as well.

With this optimization enabled, the trace looks as in Figure
\ref{fig:trace-full}. Now a lot of the language interpreter is actually gone
from the trace and what is left corresponds very closely to the loop of the
square function. The only vestige of the language interpreter is the fact that
the register list is still used to store the state of the computation. This
could be removed by some other optimization, but is maybe not really all that
bad anyway (in fact we have an experimental optimization that does exactly that,
but it is not finished).  Once we get this optimized trace, we can pass it to
the \emph{JIT backend}, which generates the correspondent machine code.

\begin{figure}
\input{code/full.txt}
\vspace{-0.4cm}
\caption{Trace when executing the Square function of Figure \ref{fig:square},
with the corresponding opcodes as comments. The constant-folding of operations
on green variables is enabled.}
\label{fig:trace-full}
\end{figure}

@@

If the JIT is enabled, things are more interesting. At the moment the JIT can
only be enabled when translating the interpreter to C, but we hope to lift that
restriction in the future. A classical tracing JIT will
interpret the program it is running until a hot loop is identified, at which
point tracing and ultimately assembler generation starts. The tracing JIT in
PyPy is operating on the language interpreter, which is written in RPython. But
RPython programs are statically translatable to C anyway. This means that interpreting the
language interpreter before a hot loop is found is clearly not desirable,
since the overhead of this double-interpretation would be significantly too big
to be practical.

What is done instead is that the language interpreter keeps running as a C
program, until a hot loop in the user program is found. To identify loops the
C version of the language interpreter is generated in such a way that at the
place that corresponds to the \texttt{can\_enter\_jit} hint profiling is
performed using the program counter of the language interpreter. Apart from this
bit of profiling, the language interpreter behaves in just the same way as
without a JIT.

When a hot user loop is identified, tracing is started. The
tracing interpreter is invoked to start tracing the language interpreter that is
running the user program. Of course the tracing interpreter cannot actually
trace the execution of the C representation of the language interpreter. Instead
it takes the state of the execution of the language interpreter and starts
tracing using a bytecode representation of the language interpreter. That means
there are two "versions" of the language interpreter embedded in the final
executable of the VM: on the one hand it is there as executable machine code, on
the other hand as bytecode for the tracing interpreter. It also means that
tracing is costly as it incurs a double interpretation overhead.

@@

\textbf{Trace Trees:} This paper ignored the problem of guards that fail often
because there are several equally likely paths through
a loop. Always falling back to interpretation in this case is not practicable.
Therefore, if we find a guard that fails often enough, we start tracing from
there and produce efficient machine code for that case, instead of always
falling back to interpretation.

@@

In this section we try to evaluate the work done so far by looking at some
benchmark numbers. Since the work is not finished, these benchmarks can only be
preliminary. Benchmarking was done on an otherwise idle machine with a 1.4
GHz Pentium M processor and 1GiB RAM, using Linux 2.6.27. All benchmarks where
run 50 times, each in a newly started process. The first run was ignored. The
final numbers were reached by computing the average of all other runs, the
confidence intervals were computed using a 95\% confidence level. All times
include the running of the tracer and machine code production to measure how
costly those are.

@@

\textbf{Benchmark 1:} The interpreter translated to C without any JIT inserted at all.

\textbf{Benchmark 2:} The tracing JIT is enabled, but no interpreter-specific
hints are applied. This corresponds to the trace in Figure
\ref{fig:trace-normal}.  The threshold when to consider a loop to be hot is 40
iterations.  As expected, this is not faster than the previous number. It is
even quite a bit slower, probably due to the overheads of the JIT, as well as
non-optimal generated machine code.

\textbf{Benchmark 3:} The hints as in Figure \ref{fig:tlr-full} are applied, which means the loop of
the square function is reflected in the trace. Constant folding of green
variables is disabled though. This corresponds to the trace in Figure
\ref{fig:trace-no-green-folding}. This by alone brings no improvement over the
previous case.

\textbf{Benchmark 4:} Same as before, but with constant folding enabled. This corresponds to the
trace in Figure \ref{fig:trace-full}. This speeds up the square function nicely,
making it about six times faster than the pure interpreter.

\textbf{Benchmark 5:} Same as before, but with the threshold set so high that the tracer is
never invoked to measure the overhead of the profiling. For this interpreter
it to be rather large, with 50\% slowdown due to profiling. This is
because the example interpreter needs to do one hash table lookup per loop
iteration. For larger interpreters (e.g. the Python one) it seems likely that
the overhead is less significant, given that many operations in Python need
hash-table lookups themselves.

\textbf{Benchmark 6:} Runs the whole computation on the tracing interpreter for estimating the
involved overheads of tracing. The trace is not actually recorded (which would be a
memory problem), so in reality the number is even higher. Due to the double
interpretation, the overhead is huge. It remains to be seen whether that will be
a problem for practical interpreters.

\textbf{Benchmark 7:} For comparison, the time of running the interpreter on top of CPython
(version 2.5.2).

\begin{figure}
\noindent
{\small
\begin{tabular}{|l|r|}
\hline
 &ratio\tabularnewline
\hline
Interpreter compiled to C, no JIT &1\tabularnewline \hline
Normal Trace Compilation &1.20\tabularnewline \hline
Unfolding of Language Interpreter Loop &XXX\tabularnewline \hline
Full Optimizations &0.17\tabularnewline \hline
Profile Overhead &1.51\tabularnewline \hline
Interpreter run by Tracing Interpreter &860.20\tabularnewline \hline
Interpreter run by CPython &256.17\tabularnewline \hline
\end{tabular}
}
\label{fig:bench1}
\caption{Benchmark results of example interpreter computing the square of
46340}
\end{figure}

XXX insert some Python benchmarks

@@

Applying a trace-based optimizer to an interpreter and adding hints to help the
tracer produce better results has been tried before in the context of the DynamoRIO
project \cite{sullivan_dynamic_2003}, which has been a great inspiration for our
work. They achieve the same unrolling of the interpreter loop so that the
unrolled version corresponds to the loops in the user program. However the
approach is greatly hindered by the fact that they trace on the machine code
level and thus have no high-level information available about the interpreter.
This makes it necessary to add quite a large number of hints, because at the
assembler level it is not really visible anymore that e.g. a bytecode string is
immutable. Also more advanced optimizations like allocation removal would
not be possible with that approach.
